Hilbert space methods to approximate Gaussian processes using Stan

### 2019-04-29 Paper/manuscript/manuscript_05.tex
fix some mistakes and make some improvements. Ready for feedback

The parameterization in \ref{approxf} is naturally in the non-centered
parameterization form with independent prior distribution on
$\beta_j$, which makes the posterior inference easier.

@@

The results from the previous section can be generalizes to a multidimensional input space with compact regular domain $\Omega=[-L_1,L_1] \times \dots \times [-L_d,L_d]$ and Dirichlet boundary conditions. 
In a $D$-dimensional input space, the total number of eigenfunctions and eigenvalues in the approximation is equal to the number of $D$-tuples, that is possible combinations of univariate eigenfunctions over all dimensions. The number $D$-tuples is given by 
%
\begin{align} \label{m_multi}
m^{\ast} = \prod_{d=1}^{D} m_d,
\end{align}

@@

@@

\noindent where $L$ describes the boundary condition, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$ is the difference between two covariate values $x$ and $x'$ in $\Omega$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are given in Equation (\ref{eigenvalue} and \ref{eigenfunction}) for the unidimensional case and in Equations (\ref{eigenvalue_multi}) and (\ref{eigenfunction_multi}) for the multidimensional case. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions is less than a predefined threshold $\varepsilon > 0$:
%
\begin{eqnarray}\label{diff_covs}
\int k(\tau) \, d\tau - 
\int \sum_{j=1}^m S_{\theta}\left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0) \, d\tau < \varepsilon.
\end{eqnarray}

The finite number $m$ of basis functions in the approximation needed to satisfy Equation (\ref{diff_covs}) depends on the non-linearity of the function to be learned, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the GP. The approximation also depends on the boundary $L$ (see Equations (\ref{eigenvalue}), (\ref{eigenfunction}), (\ref{eigenvalue}) and (\ref{eigenfunction}), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation. In the present paper, we will set $L$ an extension of the desired covariate input domain $\Psi$. Without loss of generality, we can assume $\Psi$ to be symmetric around zero, that is $\Psi=[-S,S] \subset \mathbb{R}$. We now define $L$ as
%
\begin{equation}\label{eq:boundary}
L=c \cdot S,
\end{equation}

@@

We start with an illustration on how the number of basis functions $m$ and boundary factor $c$ influences the accuracy of the HSGP approximations, separately. For this purpose, a set of noisy observations are drawn from an exact GP model with lengthscale $\ell=0.3$ and marginal variance $\alpha=1$ using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $L$ are fitted to this data. In this example, the lengthscale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictions of the estimated function and on the covariance function itself. For $c$ fixed to a large enough value, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the accuracy on the approximation and the non-linearity of the estimated function, in the sense that fewer basis functions implies inaccurate and smoother effects covariance functions (covariance functions with larger lengthscales) and consequently more linear functional forms, and conversely. 
%:Fewer basis functions imply smoother covariance functions (Paul: when would you consider a covariance function as smooth? Also, I think the main point here is that at long distances the implied covariance function becomes negative which can't be in an exact GP model.) and consequently more linear functional forms. 
Then, the higher the "wigglyness" of the function to be estimated, the more basis functions will be required. 
%(Paul: do we see this in Figure \ref{fig1_Post_J} already?). 
On the other hand, for $m$ fixed to a large enough value, Figure \ref{fig2_Post_L} shows that $c$ mainly affects the approximation near the boundaries as well as covariances at long distances.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig1_Cov_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 22mm 45mm 120mm 0mm, clip]{fig1_legend.pdf}}
\caption{(left) Posterior predictive mean functions of both regular GP model and HSGP model for different number of basis functions $m$. (right) Covariance functions of both regular GP model and HSGP model for different number of basis functions $m$.}
  \label{fig1_Post_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig2_Post_L.pdf}}
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig2_Cov_L.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 22mm 45mm 120mm 0mm, clip]{fig2_legend.pdf}}
\caption{(left) Posterior predictive mean functions of both regular GP model and HSGP model for different boundary factors $c$. (right) Covariance functions of both regular GP model and HSGP model for different boundary factors $c$.}
  \label{fig2_Post_L}
\end{figure}

@@

\begin{figure}
\begin{center}
\begin{tabular}{ c c c}
\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Post_part1.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Cov_part1.pdf} & \hspace{-6mm} \multirow{30}{1em}{ \includegraphics[scale=0.35, trim = 20mm 60mm 120mm 40mm, clip]{fig3_legend.pdf}}\\ 
\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Post_part2.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Cov_part2.pdf} &\\
\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Post_part3.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Cov_part3.pdf} &\\
\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Post_part4.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Cov_part4.pdf} & \\
\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Post_part5.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Cov_part5.pdf} &\\
\includegraphics[scale=0.25, trim = 0mm 5mm 10mm 20mm, clip]{fig3_Post_part6.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 5mm 10mm 20mm, clip]{fig3_Cov_part6.pdf} &
\end{tabular}
\end{center}
\caption{The X-axes represent the true data-generative functional lengthscales of every dataset scenario. The Y-axes represent the lengthscales estimates of GP and HSGP models. The use of different $m$ in the HSGP models is represented in separated plots. The dashed black line represents the recommended minimum lengthscales that can be closely approximate by the HSGP model in every case according to Figure \ref{fig3_lscale_vs_J_vs_c}.}
  \label{fig3_Post_part1}
\end{figure}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig6_lscale_vs_J_vs_c.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig6_lscale_vs_J_vs_c_zoomin.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ ($c = \frac{L}{S}$), and the lengthscale normalized by the half-range of the data ($\frac{\ell}{S}$). The right-side plot is a zoom in of the left-side plot.}
  \label{fig6_lscale_vs_J_vs_c}
\end{figure}

@@

Figure \ref{fig6_lscale_vs_J_vs_c} was build for a GP with a unidimensional covariance function, and results in a surface depending of three variables, $m$, $c$ and $\ell$. An equivalent figure for a GP model with a two-dimensional covariance function would result in a surface depending of four variables, $m$, $c$, $\ell_1$ and $\ell_2$, which can not be directly represented. However, as an approximation we can use the unidimensional GP conclusions in Figure \ref{fig6_lscale_vs_J_vs_c} to analyze individually the different dimensions of a multidimensional GP model.

@@

For this analysis, we will use various datasets consisting of noisy draws from a GP prior model with a squared exponential covariance function (Paul: Did we talk about the applied kernels previously in Section 3? Gabi: I've mentioned it in line 491) and varying lengthscale values. Different values of the number of basis functions $m$ are used when estimating the HSGP models, and the boundary factor $c$ is set to a valid and optimum value in every case.

Figure \ref{fig7_posterior_varing_lscale_part1} shows the posterior predictions of the regular GP and HSGP models fitted to those datasets. The Lengthscale estimates as obtained by regular GP and HSGP models are depicted in Figure \ref{fig8_Tlscale_vs_Elscale}. Further, Figure \ref{fig9_MSE_varing_lscale} shows the root mean square error (RMSE) of the HSGP models against the regular GP models as a function of the lengthscale and number of basis functions $m$.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.4, trim = 10mm 95mm 5mm 30mm, clip]{fig7_legend.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part9.pdf}}\\
\vspace{-3mm}
\caption{Posterior predictive means of both regular GP and HSGP models, fitted over different datasets simulated from square exponential GP models with different characteristic lengthscales ($lscale$) and same marginal variance ($\alpha$) as the data-generative function (\textit{True function}).}
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

\begin{figure}
\begin{flushleft}
\begin{tabular}{ c c c c}
\includegraphics[scale=0.28, trim = 0mm 30mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part1.pdf} \vspace{0mm} & \hspace{-5mm} \includegraphics[scale=0.28, trim = 22mm 30mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part2.pdf} \vspace{0mm} & \hspace{-5mm} \includegraphics[scale=0.28, trim = 22mm 30mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part3.pdf} \vspace{0mm} & \hspace{-8mm} \multirow{3}{1em}{\vspace{-9mm} \includegraphics[scale=0.35, trim = 10mm 105mm 130mm 45mm, clip]{fig8_legend.pdf}}\\ 
\includegraphics[scale=0.28, trim = 0mm 10mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part4.pdf} \vspace{0mm}  & \hspace{-5mm} \includegraphics[scale=0.28, trim = 22mm 10mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part5.pdf} \vspace{0mm} & \hspace{-5mm} \includegraphics[scale=0.28, trim = 22mm 10mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part6.pdf} \vspace{0mm}
\end{tabular}
\end{flushleft}
\caption{The X-axes represent the true data-generative functional lengthscales of every dataset scenario. The Y-axes represent the lengthscales estimates of GP and HSGP models. The use of different $m$ in the HSGP models is represented in separated plots. The dashed black line represents the recommended minimum lengthscales that can be closely approximate by the HSGP model in every case according to Figure \ref{fig6_lscale_vs_J_vs_c}.}
  \label{fig8_Tlscale_vs_Elscale}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=0.35]{fig9_MSE_varing_lscale.pdf}
\caption{RMSE of the HSGP models for the different datasets with different wiggly effects (lengthscales) and number of basis functions $m$.}
  \label{fig9_MSE_varing_lscale}
\end{figure}

We can appreciate an agreement between Figures \ref{fig8_Tlscale_vs_Elscale} and \ref{fig9_MSE_varing_lscale} except for small lengthscales and long lengthscales. We observe small RMSE for large lengthscales, however we appreciate mismatch between large lengthscale estimates of GP and HSGP models in Figure \ref{fig8_Tlscale_vs_Elscale}. This is because big changes on large lengthscales, which might be clearly appreciated in Figure \ref{fig8_Tlscale_vs_Elscale}, might result in similar posterior and therefore a small RMSE. On the other side, we observe high RMSE for small lengthscales, however we appreciate match between small lengthscale estimates of GP and HSGP models in Figure \ref{fig8_Tlscale_vs_Elscale}. This is because very small changes on small lengthscales, which might not be appreciated in Figure \ref{fig8_Tlscale_vs_Elscale}, can have big effects on the posteriors and therefore high RMSE.

@@

\noindent $L$ is the boundary and $m$ the number of basis functions. The parameters $\beta_j$ are $\mathcal{N}(0,1)$ distributed, and $\alpha$ and $\ell$ are the marginal variance and lengthscale parameters of the approximate covariance function, respectively.

@@

Figure \ref{fig10_Posteriors_exI} shows the posteriors distributions of the three models, the regular GP, the HSGP with $m=80$ basis functions and boundary factor $c=1.2$ ($L=c\cdot 1= 1.2$ (\ref{eq:boundary})), and the splines model with 80 knots. The true data-generative function and the noisy observations are also plotted in the Figure. The right and left extremes and three small regions in the middle of the plot correspond to out-of-sample or test data which have not been taking part on training the models. The test data have been plotted as crosses and the training data as circles. The test data located at the extremes of the plot are used for assessing model extrapolation, and the test data located in the middle are used for assessing model interpolation.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{fig10_Posteriors_exI.pdf}
\caption{Posterior distributions of the proposed HSGP model, the regular GP model, and the Splines model.}
  \label{fig10_Posteriors_exI}
\end{figure}

In order to assess performance of the models as a function the number of basis functions and number of knots, different models with different number of basis functions for the HSGP model and different number of knots for the splines model have been fitted. Figure \ref{fig11_MSE_exI_inter} shows the  standardized root mean squared error (SRMSE), for interpolation and extrapolating data, as a function of the number of basis functions and knots. The SRMSE is computed against the true function. Model interpolation has been assessed either using the test data for interpolating and the training data. From Figures \ref{fig10_Posteriors_exI} and \ref{fig11_MSE_exI_inter}, it can be seen a close approximation of the HSGP model to the regular GP model, for interpolating and extrapolating data. However, the splines model does not extrapolate data properly. For interpolation both models show roughly similar performance.

>>>>

The parameterization in (\ref{approxf}) is naturally in the non-centered
parameterization form with independent prior distribution on
$\beta_j$, which makes the posterior inference easier.

@@

The results from the previous section can be generalizes to a multidimensional input space with compact regular domain $\Omega=[-L_1,L_1] \times \dots \times [-L_d,L_d]$ and Dirichlet boundary conditions. 
In a $D$-dimensional input space, the total number of eigenfunctions and eigenvalues in the approximation is equal to the number of $D$-tuples, that is possible combinations of univariate eigenfunctions over all dimensions. The number of $D$-tuples is given by 
%
\begin{align} \label{m_multi}
m^{\ast} = \prod_{d=1}^{D} m_d,
\end{align}

@@

(Gabi: The following paragraph is a first idea to introduce that the recommendations and diagnostics depends on the kernel considered. And that, at the moment, we have built these recommendations only for the square exponential kernel. I know that what I wrote in not enough and should be a bit more elaborate; your suggestions where and how elaborate it would be useful.)

Ultimately, these recommendations lie on the relationships among the number of basis functions, the boundary factor and the lengthscale of the function, which depend on the particular choice of the kernel function. In this work we built these relationships for the square exponential covariance function. For other kernels, the relationships will be slightly different.

@@

\noindent where $L$ describes the boundary condition, $j$ is the index for the eigenfunctions and eigenvalues, and $\tau=x-x'$ is the difference between two covariate values $x$ and $x'$ in $\Omega$. The eigenvalues $\lambda_j$ and eigenfunctions $\phi_j$ are given in Equations (\ref{eigenvalue}) and (\ref{eigenfunction}) for the unidimensional case and in Equations (\ref{eigenvalue_multi}) and (\ref{eigenfunction_multi}) for the multidimensional case. The number of basis functions can be truncated at some finite positive value $m$ such that the difference between the densities of the exact and approximate covariance functions is less than a predefined threshold $\varepsilon > 0$:
%
\begin{eqnarray}\label{diff_covs}
\int k(\tau) \, d\tau - 
\int \sum_{j=1}^m S_{\theta}\left(\sqrt{\lambda_j} \right) \phi_j(\tau) \phi_j(0) \, d\tau < \varepsilon.
\end{eqnarray}

The finite number $m$ of basis functions in the approximation needed to satisfy Equation (\ref{diff_covs}) depends on the non-linearity of the function to be learned, that is on its lengthscale $\ell$, which constitutes a hyperparameter of the GP. The approximation also depends on the boundary $L$ (see Equations (\ref{eigenvalue}), (\ref{eigenfunction}), (\ref{eigenvalue_multi}) and (\ref{eigenfunction_multi})), which will affect its accuracy especially near the boundaries. As we will see later on, $L$ will also influence the number of basis functions required in the approximation. In the present paper, we will set $L$ an extension of the desired covariate input domain $\Psi$. Without loss of generality, we can assume $\Psi$ to be symmetric around zero, that is $\Psi=[-S,S] \subset \mathbb{R}$. We now define $L$ as
%
\begin{equation}\label{eq:boundary}
L=c \cdot S,
\end{equation}

@@

We start with an illustration on how the number of basis functions $m$ and boundary factor $c$ influences the accuracy of the HSGP approximations, separately. For this purpose, a set of noisy observations are drawn from an exact GP model with lengthscale $\ell=0.3$ and marginal variance $\alpha=1$, using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $L$ are fitted to this data. In this example, the lengthscale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures \ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictions of the estimated function and on the covariance function itself. For $c$ fixed to a large enough value, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the accuracy on the approximation and the non-linearity of the estimated function, in the sense that fewer basis functions implies inaccurate and smoother effects covariance functions (covariance functions with larger lengthscales) and consequently more linear functional forms, and conversely. 
%:Fewer basis functions imply smoother covariance functions (Paul: when would you consider a covariance function as smooth? Also, I think the main point here is that at long distances the implied covariance function becomes negative which can't be in an exact GP model.) and consequently more linear functional forms. 
Then, the higher the "wigglyness" of the function to be estimated, the more basis functions will be required. 
%(Paul: do we see this in Figure \ref{fig1_Post_J} already?). 
On the other hand, for $m$ fixed to a large enough value, Figure \ref{fig2_Post_L} shows that $c$ mainly affects the approximation near the boundaries as well as covariances at long distances.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig1_Post_J.pdf}}
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig1_Cov_J.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 22mm 45mm 120mm 0mm, clip]{fig1_legend.pdf}}
\caption{Mean posterior predictive functions (left) and covariance functions (right) of both the regular GP model and the HSGP model for different number of basis functions $m$, with the boundary factor fixed to a large enough value.}
  \label{fig1_Post_J}
\end{figure}

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig2_Post_L.pdf}}
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig2_Cov_L.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 22mm 45mm 120mm 0mm, clip]{fig2_legend.pdf}}
\caption{Mean posterior predictive functions (left) and covariance functions (right) of both the regular GP model and the HSGP model for different values of the boundary factor $c$, with a large enough fixed number of basis functions.}
  \label{fig2_Post_L}
\end{figure}

@@

\begin{figure}
\begin{center}
\begin{tabular}{ c c c}
\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Post_part1.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Cov_part1.pdf} & \hspace{-6mm} \multirow{30}{1em}{ \includegraphics[scale=0.35, trim = 20mm 60mm 120mm 40mm, clip]{fig3_legend.pdf}}\\ 
\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Post_part2.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Cov_part2.pdf} &\\
\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Post_part3.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Cov_part3.pdf} &\\
\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Post_part4.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Cov_part4.pdf} & \\
\includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Post_part5.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 30mm 10mm 20mm, clip]{fig3_Cov_part5.pdf} &\\
\includegraphics[scale=0.25, trim = 0mm 5mm 10mm 20mm, clip]{fig3_Post_part6.pdf} & \hspace{-5mm} \includegraphics[scale=0.25, trim = 0mm 5mm 10mm 20mm, clip]{fig3_Cov_part6.pdf} &
\end{tabular}
\end{center}
\caption{Mean posterior predictive functions (left) and covariance functions (right) of both the regular GP model and the HSGP model for different number of basis functions $m$ and for different values of the boundary factor $c$.}
  \label{fig3_Post_part1}
\end{figure}

@@

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig6_lscale_vs_J_vs_c.pdf}}
\hspace{3mm}
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig6_lscale_vs_J_vs_c_zoomin.pdf}}
\caption{Relation among the minimum number of basis functions $m$, the boundary factor $c$ ($c = \frac{L}{S}$) and the lengthscale normalized by the half-range of the data ($\frac{\ell}{S}$), in the case of a square exponential covariance function. The right-side plot is a zoom in of the left-side plot.}
  \label{fig6_lscale_vs_J_vs_c}
\end{figure}

@@

Figure \ref{fig6_lscale_vs_J_vs_c} was build for a GP with a unidimensional covariance function, and results in a surface depending on three variables, $m$, $c$ and $\ell$. An equivalent figure for a GP model with a two-dimensional covariance function would result in a surface depending on four variables, $m$, $c$, $\ell_1$ and $\ell_2$, which can not be directly represented. However, as an approximation we can use the unidimensional GP conclusions in Figure \ref{fig6_lscale_vs_J_vs_c} to analyze individually the different dimensions of a multidimensional GP model.

(Gabi: The following paragraph is another idea to introduce that the diagnosis plot in Figure \ref{fig6_lscale_vs_J_vs_c} depends on the kernel considered. At the moment we have built this figure only for the square exponential kernel.)

Ultimately, Figure \ref{fig6_lscale_vs_J_vs_c} showing the relationships among the number of basis functions, the boundary factor and the lengthscale of the function depends on the particular choice of covariance function. In this work we have built these relationships only for the square exponential covariance function. For other kernels, the Figure will be slightly different.

@@

For this analysis, we will use various datasets consisting of noisy draws from a GP prior model with a squared exponential covariance function (Paul: Did we talk about the applied kernels previously in Section 3? Gabi: I did some attempt about that in previous subsection, but I am not sure whether succesfully.) and varying lengthscale values. Different values of the number of basis functions $m$ are used when estimating the HSGP models, and the boundary factor $c$ is set to a valid and optimum value in every case.

Figure \ref{fig7_posterior_varing_lscale_part1} shows the posterior predictions of both regular GP and HSGP models fitted to those datasets. The lengthscale estimates as obtained by regular GP and HSGP models are depicted in Figure \ref{fig8_Tlscale_vs_Elscale}. As commented previously, an accurate estimate of the lengthscale can be a good indicator of a close approximation of the HSGP model to the regular GP model. Further, Figure \ref{fig9_MSE_varing_lscale} shows the root mean square error (RMSE) of the HSGP models, computed against the regular GP models as a function of the lengthscale and number of basis functions $m$.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.4, trim = 10mm 95mm 5mm 30mm, clip]{fig7_legend.pdf}}\\
\vspace{-2mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part9.pdf}}\\
\vspace{-3mm}
\caption{Mean posterior predictions of both regular GP and HSGP models, fitted over various datasets drawn from square exponential GP models with different characteristic lengthscales ($lscale$) and same marginal variance ($\alpha$) as the data-generating functions (\textit{True function}).}
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

\begin{figure}
\begin{flushleft}
\begin{tabular}{ c c c c}
\includegraphics[scale=0.28, trim = 0mm 30mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part1.pdf} \vspace{0mm} & \hspace{-5mm} \includegraphics[scale=0.28, trim = 22mm 30mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part2.pdf} \vspace{0mm} & \hspace{-5mm} \includegraphics[scale=0.28, trim = 22mm 30mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part3.pdf} \vspace{0mm} & \hspace{-8mm} \multirow{3}{1em}{\vspace{-9mm} \includegraphics[scale=0.35, trim = 10mm 105mm 130mm 45mm, clip]{fig8_legend.pdf}}\\ 
\includegraphics[scale=0.28, trim = 0mm 10mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part4.pdf} \vspace{0mm}  & \hspace{-5mm} \includegraphics[scale=0.28, trim = 22mm 10mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part5.pdf} \vspace{0mm} & \hspace{-5mm} \includegraphics[scale=0.28, trim = 22mm 10mm 10mm 20mm, clip]{fig8_Tlscale_vs_Elscale_part6.pdf} \vspace{0mm}
\end{tabular}
\end{flushleft}
\caption{Data-generating functional lengthscales (X-axis), of the various datasets illustrated in Figure \ref{fig7_posterior_varing_lscale_part1}, versus the corresponding lengthscale estimates from the regular GP and HSGP models (Y-axis). 95\% confident intervals of the lengthscale estimates are plotted as dot lines. The different plots represent the use of different number of basis functions $m$ in the HSGP model. The dashed black line represents the recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c} that can be closely approximated by the HSGP model in every case.}
  \label{fig8_Tlscale_vs_Elscale}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=0.35]{fig9_MSE_varing_lscale.pdf}
\caption{RMSE of the HSGP models with different number of basis functions $m$, for the various datasets with different wiggly effects (lengthscales).}
  \label{fig9_MSE_varing_lscale}
\end{figure}

Comparing the information of models approximation provided by Figures \ref{fig8_Tlscale_vs_Elscale} and \ref{fig9_MSE_varing_lscale}, it can be appreciated an agreement between both figures, excepting for small lengthscales and long lengthscales. We observe small RMSE for large lengthscales in Figure \ref{fig9_MSE_varing_lscale}, however there exist mismatch between large lengthscale estimates of GP and HSGP models in Figure \ref{fig8_Tlscale_vs_Elscale}. This is because big changes on large lengthscales, which might be clearly appreciated in Figure \ref{fig8_Tlscale_vs_Elscale}, might result in similar posterior predictions and therefore a small RMSE. On the other side, we observe high RMSE for small lengthscales in Figure \ref{fig9_MSE_varing_lscale}, however there exist match between small lengthscale estimates of GP and HSGP models in Figure \ref{fig8_Tlscale_vs_Elscale}. This is because very small changes on small lengthscales, which might not be appreciated in Figure \ref{fig8_Tlscale_vs_Elscale}, can have big effects on the posteriors predictions and therefore high RMSE.

@@

\noindent In the previous equations, $L$ is the boundary and $m$ the number of basis functions. The parameters $\beta_j$ are $\mathcal{N}(0,1)$ distributed, and $\alpha$ and $\ell$ are the marginal variance and lengthscale parameters, respectively, of the approximate covariance function.

@@

Figure \ref{fig10_Posteriors_exI} shows the posteriors predictive distributions of the three models, the regular GP, the HSGP with $m=80$ basis functions and boundary factor $c=1.2$ ($L=c\cdot 1= 1.2$ (\ref{eq:boundary})), and the splines model with 80 knots. The true data-generative function and the noisy observations are also plotted in the Figure; the sample observations are plotted as circles and the out-of-sample or test data, which have not been taking part on training the models, are plotted as crosses. The test data located at the extremes of the plot are used for assessing model extrapolation, and the test data located in the middle are used for assessing model interpolation.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{fig10_Posteriors_exI.pdf}
\caption{Posterior predictive distributions of the proposed HSGP model, the regular GP model, and the Splines model. 95\% credible intervals are plotted as dot lines.}
  \label{fig10_Posteriors_exI}
\end{figure}

In order to assess performance of the models as a function the number of basis functions and number of knots, different models with different number of basis functions for the HSGP model and different number of knots for the splines model have been fitted. Figure \ref{fig11_MSE_exI_inter} shows the standardized root mean squared error (SRMSE), for interpolation and extrapolating data, as a function of the number of basis functions and knots. The SRMSE is computed against the data-generating function. From Figures \ref{fig10_Posteriors_exI} and \ref{fig11_MSE_exI_inter}, it can be seen a close approximation of the HSGP model to the regular GP model, for interpolating and extrapolating data. However, the splines model does not extrapolate data properly. Both models show roughly similar interpolating performance.

Figure \ref{fig11_time_exI} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the splines model. The computational time is represented in the Y-axis of the figure, which is in a logarithmic scale. The HSGP model is on average roughly 400 times faster than the regular GP, in the particular case of applying over this dataset.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig11_MSE_exI_inter.pdf}}
\subfigure{\includegraphics[scale=0.35, trim = 0mm 0mm 10mm 0mm, clip]{fig11_MSE_exI_extra.pdf}}
\subfigure{\includegraphics[scale=0.40, trim = 22mm 60mm 120mm 0mm, clip]{fig11_MSE_exI_legend.pdf}}
\caption{Standarized root mean square error (SRMSE) of the different methods against the data-generating function. (left) SRMSE for interpolation. (right) SRMSE for extrapolation. The Standard deviation of the mean of the SRMSE are plotted as dot lines.}
  \label{fig11_MSE_exI_inter}
\end{figure}
